*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-23 Code/Manuscript_O2.Rmd
Fixed citations in intro

*P. marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures (Rocap et al., 2003). The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers (2018, Partensky et al. 2018).

@@

Current niches of *P. marinus* span ocean water columns
[@chisholmNovelFreelivingProchlorophyte1988; @veldhuisVerticalDistributionPigment1990; @partenskyProchlorococcusMarinePhotosynthetic1999] and extend into regions with varying dissolved oxygen concentrations
[@partenskyProchlorococcusAdvantagesLimits2010; @goerickeNovelNicheProchlorococcus2000] and even into anoxic oceans [@lavinNovelLineagesProchlorococcus2010], [@ulloaCyanobacteriumProchlorococcusHas2021; @garcia-robledoCrypticOxygenCycling2017].

Our changing climate, characterized by warming ocean waters, is rapidly altering conditions for these specialized clades of marine picophytoplankton. Predictions indicate a net global increase of *P. marinus* cell abundances of 29%, along with  poleward latitudinal shifts of at least 10° in marine phytoplankton niches by the end of this century [@flombaumPresentFutureGlobal2013][@bartonAnthropogenicClimateChange2016] in response to warming waters, with increases in *P. marinus* of approximately 50% in the more poleward regions of their distributions.

Near the equator, photoperiod remains relatively constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attentuate with depth as dawn and dusk light at depth drops below the level needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches @liDiatomGrowthResponses2017 @prezelinDielPeriodicityPhytoplankton1992. For example, @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.

@@

We used the OceansMap Protein Portal (OPP; https://proteinportal.whoi.edu/) to analyze the distribution of proteins from clades of *P. marinus* in samples tanken across a range of [O~2~] and depth, which in turn correlates to peak light at the site of sampling. We therefore analyzed the growth and physiological responses of representative strains from three clades ofr *P. marinus* under a matrix of [O~2~], light levels, spectral waveband ranges, and photoperiods, to approximate eco-physiological conditions representative of current and hypothetical future ocean zones. *P. marinus* MED4, a clade HLI strain, was isolated near the ocean surface (5m depth) of the Mediterranean Sea where [O~2~] is near saturation, light levels are high and spectral bias from full solar irradiance is minimal. *P. marinus* SS120, a clade LLIII strain, was isolated from the Sargasso Sea at a depth of 120 m, while *P. marinus* MIT9313, a clade LLIV strain, was isolated from the North Atlantic Gulf Stream at a depth of 135 m. At these depths, light attenuation and spectral shifts occur, resulting in low blue and green light, while [O~2~] varies from near-surface saturation levels to decreased concentrations, but does not necessarily decrease systematically with depth [@billheimerOxygenSeasonalityUtilization2021].

>>>>

*P. marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures [@rocapGenomeDivergenceTwo2003]. The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m [@partenskyProchlorococcusMarinePhotosynthetic1999]. LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017; @johnsonEnergeticsGrowthKinetics1999]. *Prochlorococcus marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers [@partenskyComparisonPhotosyntheticPerformances2018].

@@

Current niches of *P. marinus* span ocean water columns
[ @veldhuisVerticalDistributionPigment1990; @chisholmNovelFreelivingProchlorophyte1988; @partenskyProchlorococcusMarinePhotosynthetic1999] and extend into regions with varying dissolved oxygen concentrations
[@partenskyProchlorococcusAdvantagesLimits2010; @goerickeNovelNicheProchlorococcus2000] and even into anoxic oceans [@lavinNovelLineagesProchlorococcus2010, @ulloaCyanobacteriumProchlorococcusHas2021; @garcia-robledoCrypticOxygenCycling2017].

Our changing climate, characterized by warming ocean waters, is rapidly altering conditions for these specialized clades of marine picophytoplankton. Predictions indicate a net global increase of *P. marinus* cell abundances of 29%, along with  poleward latitudinal shifts of at least 10° in marine phytoplankton niches by the end of this century [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016] in response to warming waters, with increases in *P. marinus* of approximately 50% in the more poleward regions of their distributions.

Near the equator, photoperiod remains relatively constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attentuate with depth as dawn and dusk light at depth drops below the level needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches [@liDiatomGrowthResponses2017; @prezelinDielPeriodicityPhytoplankton1992]. For example, Vaulot et al. @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.

@@

We used the OceansMap Protein Portal (OPP; https://proteinportal.whoi.edu/) to analyze the distribution of proteins from clades of *P. marinus* in samples tanken across a range of [O~2~] and depth, which in turn correlates to peak light at the site of sampling. We therefore analyzed the growth and physiological responses of representative strains from three clades of *P. marinus* under a matrix of [O~2~], light levels, spectral waveband ranges, and photoperiods, to approximate eco-physiological conditions representative of current and hypothetical future ocean zones. *P. marinus* MED4, a clade HLI strain, was isolated near the ocean surface (5m depth) of the Mediterranean Sea where [O~2~] is near saturation, light levels are high and spectral bias from full solar irradiance is minimal. *P. marinus* SS120, a clade LLIII strain, was isolated from the Sargasso Sea at a depth of 120 m, while *P. marinus* MIT9313, a clade LLIV strain, was isolated from the North Atlantic Gulf Stream at a depth of 135 m. At these depths, light attenuation and spectral shifts occur, resulting in low blue and green light, while [O~2~] varies from near-surface saturation levels to decreased concentrations, but does not necessarily decrease systematically with depth [@billheimerOxygenSeasonalityUtilization2021].
